Relationships in the extremes and their influence on competition and coexistence: Supporting information

### 2021-08-04 paper/Paper.Rnw

%***DAN: stopped here 2021 07 20
%[As we know], anthropogenic climate change threatens almost all aspects of life including via extreme climatic events, which have only become more extreme - increasing in magnitude and frequency - and have contributed to both the increase in global means as well as an increase in environmental variability (citation). 
%There are many studies on the influence of environmental extremes and variability on competition and coexistence \citep{Adler_2006,  Parepa_2013, Diez_2012}; however, the study of the influence of relationships and interactions between drivers in their extremes on competition and coexistence, is lacking in the scientific literature.
%;we review these studies here  In one of the first empirical tests for storage effect and its requirements, \cite{Adler_2006} showed that climate variability - including extreme events - promoted the coexistence of grass species in Kansas prairie, specifically their ability to recover from low densities. Moreover, in studies on the invasibility of plants, the density and competitive balance of invasive Japanese knotweed increases as resource and thus climate variability increased.\citep{Parepa_2013} 
Environmental fluctuations are becoming more influenced by extreme climatic events (Hansen et al 2012) as extreme events increase in frequency, duration, and intensity \citep{Meehl_2004, Ummenhofer_2017}. Extreme events (heat-waves, floods, hurricanes, etc.) alter the means and variation of environmental fluctuations, and thus have implications for coexistence and competition. Competitive output and community composition of grassland plants has been shown to be impacted by heat-waves and extreme drought \citep{White_2000, White_2001, Hoover_2014}. Studies on the direct effect of extreme events on competing species are slim (but see above); however, the consequences of change in environmental means, which may be due to extreme events or gradual climate change in general, on coexistence is more commonly studied.  \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as rainfall levels are forecasted to decrease. Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. More variable environmental fluctuations are even less so considered compared to changes in spatial means; although, environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}.

@@

%If extreme events are more temporally extensive/expansive than regular environmental fluctuations, extreme events may cause the environmental response of two species who depend on temporally separated indicators of growth to become more correlated in the extremes. 
To demonstrate plausibility of ATA altering coexistence between species, we carried out long time scale simulations using a simple lottery model. 
To flesh out the idea of how ATAs in environmental response might manifest and modfiy the outcome of coexistence, we introduce an example of a two species grass-forb plant community (inspired by \cite{Chesson_2012}. The community follows the criteria required for temporal storage effects to operate. The forb dominates and has high fecundity when a sufficient threhold of rain is reached in the early season. The grass dominates and has high fecundity when there is sufficeint rain in the late season. Competition is influenced by the resource comsumption of growing juvenile plants. Dormant seeds are unaffected by competiton and unfavorable environment and can persist in the community. When rain conditions fluctuate from year to year, relative abundance and dominance of the grass-forb community also fluctuate year to year. This produces a relationship between the grass's and forb's environmental responses (fecundity) that is symmetrically associated. ATAs in environmental response can be caused by extreme climatic events. If the community is affected by frequent and long drought with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the lower values or left tail associated. Conversely, if the community is effected by frequent and long floods with some years of normalcy, the grass-forb environmental relationship may become more correlated in the higher values or right-tail associated. In Figure 2 we see a possible outcome of ATAs between environmental responses. In simulating their populations,  ATAs can cause species to have longer periods of extreme dominance and lower growth rates overall.      
%move this paragraph to discussion
Previously, storage effects were described as the contribution of covariance between the environment and competition to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019}. 
However, the description of storage effects as the covariance between drivers is a (potentially dangerous) misnomer in terminology, and we argue that storage effects, instead, should be described as the contribution of the relationship between drivers. 
All aspects and contributing values of the relationship between environmental response and competitive pressure are not fully captured by only the covariance between them because a measure of covariance does not capture the strucutre of associativity throuhgout the distribution of the relationship and any asymmetries there may be. 
Thus, aligning with the decomposition presented in \cite{Ellner_2019}, we propose storage effect can be broken down further into the contribution of ATA [to storage effect/coexistence] and the contribution of symmetric association [to storage effect/coexistence].

>>>>

%***DAN: stopped here 2021 07 20

%[As we know], anthropogenic climate change threatens almost all aspects of life including via extreme climatic events, which have only become more extreme - increasing in magnitude and frequency - and have contributed to both the increase in global means as well as an increase in environmental variability (citation). 
%There are many studies on the influence of environmental extremes and variability on competition and coexistence \citep{Adler_2006,  Parepa_2013, Diez_2012}; however, the study of the influence of relationships and interactions between drivers in their extremes on competition and coexistence, is lacking in the scientific literature.
%;we review these studies here  In one of the first empirical tests for storage effect and its requirements, \cite{Adler_2006} showed that climate variability - including extreme events - promoted the coexistence of grass species in Kansas prairie, specifically their ability to recover from low densities. Moreover, in studies on the invasibility of plants, the density and competitive balance of invasive Japanese knotweed increases as resource and thus climate variability increased.\citep{Parepa_2013} 
Environmental fluctuations are becoming more influenced by extreme climatic events (Hansen et al 2012) as extreme events increase in frequency, duration, and intensity \citep{Meehl_2004, Ummenhofer_2017}. Extreme events (heat-waves, floods, hurricanes, etc.) alter the means and variation of environmental fluctuations, and thus have implications for coexistence and competition. Competitive output and community composition of grassland plants has been shown to be impacted by heat-waves and extreme drought \citep{White_2000, White_2001, Hoover_2014}. Studies on the direct effect of extreme events on competing species are slim (but see above); however, the consequences of change in environmental means, which may be due to extreme events or gradual climate change in general, on coexistence is more commonly studied.  \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as rainfall levels are forecasted to decrease. Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. More variable environmental fluctuations are even less so considered compared to changes in spatial means; although, environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}.

@@

%If extreme events are more temporally extensive/expansive than regular environmental fluctuations, extreme events may cause the environmental response of two species who depend on temporally separated indicators of growth to become more correlated in the extremes. 
Extreme events can have important ecological impacts for controphic competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive (last longer, have higher duration) than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (similar, correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. This is a change in ATA not just from the event alone, but from the relationship between species. Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition, they do not directly consider ATA between species response caused by environment.
To demonstrate plausibility of ATA altering coexistence between species, we carried out long time scale simulations using a simple lottery model. 
To flesh out the idea of how ATAs in environmental response might manifest and modfiy the outcome of coexistence, we introduce an example of a two species grass-forb plant community (inspired by \cite{Chesson_2012}. The community follows the criteria required for temporal storage effects to operate. The forb dominates and has high fecundity when a sufficient threhold of rain is reached in the early season. The grass dominates and has high fecundity when there is sufficeint rain in the late season. Competition is influenced by the resource comsumption of growing juvenile plants. Dormant seeds are unaffected by competiton and unfavorable environment and can persist in the community. When rain conditions fluctuate from year to year, relative abundance and dominance of the grass-forb community also fluctuate year to year. This produces a relationship between the grass's and forb's environmental responses (fecundity) that is symmetrically associated. ATAs in environmental response can be caused by extreme climatic events. If the community is affected by frequent and long drought with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the lower values or left tail associated. Conversely, if the community is effected by frequent and long floods with some years of normalcy, the grass-forb environmental relationship may become more correlated in the higher values or right-tail associated. In Figure 2 we see a possible outcome of ATAs between environmental responses. In simulating their populations,  ATAs can cause species to have longer periods of extreme dominance and lower growth rates overall.

%move this paragraph to discussion
Previously, storage effects were described as the contribution of covariance between the environment and competition to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019}. 
However, the description of storage effects as the covariance between drivers is a (potentially dangerous) misnomer in terminology, and we argue that storage effects, instead, should be described as the contribution of the relationship between drivers. 
All aspects and contributing values of the relationship between environmental response and competitive pressure are not fully captured by only the covariance between them because a measure of covariance does not capture the strucutre of associativity throuhgout the distribution of the relationship and any asymmetries there may be. 
Thus, aligning with the decomposition presented in \cite{Ellner_2019}, we propose storage effect can be broken down further into the contribution of ATA [to storage effect/coexistence] and the contribution of symmetric association [to storage effect/coexistence].

### 2021-08-04 paper/SupMat.Rnw
worked on editing the Intro

%The following is where you load in the numeric results that will be embedded in the text
<<eval=T,echo=F,message=F,warning=F>>=
system(paste("biber", sub("\\.Rnw$", "", current_input())))

#place R code here for loading in necessary variables from Results
#ordinarily this would be a readRDS command, but I'm just defining a variable here for now
allregres<-matrix(rnorm(10),2,5)
@

@@

Next, proceed as follows to get estimates of the terms in the decomposition (\ref{eq:decomp_overall_rjbsi}). 
\begin{enumerate}
\item Use $\epsilon_i^0=0$, as in (\ref{eq:decomp:baseline_j}), and this is the exact value.
\item Estimate $\olep_j^E$, following (\ref{eq:decomp:contrib_of_var_E_j}), as
$\widehat{\olep_j^E}=\mean_k[\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]-\epsilon_j^0$, an estimate
with standard error $\se \left( \widehat{\olep_j^E} \right) = \sd_k [\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]/\sqrt{M}$.
\item Estimate $\olep_j^C$, following (\ref{eq:decomp:contrib_of_var_C_j}), as
$\widehat{\olep_j^C}=\mean_k[\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]-\epsilon_j^0$, which has 
standard error $\se \left( \widehat{\olep_j^C} \right) = \sd_k [\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Note that $\tilde{b}_j$ and $\tilde{b}_j^\shp$ in (\ref{eq:decomp:j:EshpC}) are standard normally distributed and
independent, so $\tilde{b}_j^\shp - \tilde{b}_j$ is normally distributed with mean $0$ and variance $2$. So we can 
estimate $\olep_j^{(E \shp C)}$, following (\ref{eq:decomp:j:EshpC}), as 
$\widehat{\olep_j^{(E \shp C)}} = \mean_k [\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))]-\widehat{\olep_j^E}-\widehat{\olep_j^C}-\epsilon_j^0$.
The standard error of this expression is $\se \left(  \widehat{\olep_j^{(E \shp C)}} \right) = 
\sd_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))-\ln(1-\delta+\delta\exp(u^{(k)}-\sigma^2/2))-
\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Estimate $\olep_j^{[EC]}$, following (\ref{eq:decomp:j:EChard}), as $\widehat{\olep_j^{[EC]}}=\mean_k [-\ln(1-\delta+\delta \exp(\sigma \sqrt{2} u^{(k)}))],$ which has $\se \left( \widehat{\olep_j^{[EC]}} \right) = \sd_k[-\ln(1-\delta+\delta \exp(\sigma \sqrt{2} u^{(k)}))]/\sqrt{M}$.
\item Use $\olep_j^{[E \nat C]}=0$, following (\ref{eq:decomp:j:EnatC}). This is exact.
\item Finally, use $\olr_{j \bs i} = 0$, again exact. 
\end{enumerate}
As for (\ref{eq:decomp_overall_IGR}), it is again easy to see that (\ref{eq:decomp_overall_rjbsi}) applies for the 
estimates as well as for the quantities being estimated.

>>>>

%The following is where you load in the numeric results that will be embedded in the text
%<<eval=T,echo=F,message=F,warning=F>>=
%system(paste("biber", sub("\\.Rnw$", "", current_input())))

%#place R code here for loading in necessary variables from Results
%#ordinarily this would be a readRDS command, but I'm just defining a variable here for now
%allregres<-matrix(rnorm(10),2,5)
%@

@@

Next, proceed as follows to get estimates of the terms in the decomposition (\ref{eq:decomp_overall_rjbsi}). 
\begin{enumerate}
\item Use $\epsilon_j^0=0$, as in (\ref{eq:decomp:baseline_j}), and this is the exact value.
\item Estimate $\olep_j^E$, following (\ref{eq:decomp:contrib_of_var_E_j}), as
$\widehat{\olep_j^E}=\mean_k[\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]-\epsilon_j^0$, an estimate
with standard error $\se \left( \widehat{\olep_j^E} \right) = \sd_k [\ln(1-\delta+\delta\exp(\sigma u^{(k)}-\sigma^2/2))]/\sqrt{M}$.
\item Estimate $\olep_j^C$, following (\ref{eq:decomp:contrib_of_var_C_j}), as
$\widehat{\olep_j^C}=\mean_k[\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]-\epsilon_j^0$, which has 
standard error $\se \left( \widehat{\olep_j^C} \right) = \sd_k [\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Note that $\tilde{b}_j$ and $\tilde{b}_j^\shp$ in (\ref{eq:decomp:j:EshpC}) are standard normally distributed and
independent, so $\tilde{b}_j^\shp - \tilde{b}_j$ is normally distributed with mean $0$ and variance $2$. So we can 
estimate $\olep_j^{(E \shp C)}$, following (\ref{eq:decomp:j:EshpC}), as 
$\widehat{\olep_j^{(E \shp C)}} = \mean_k [\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))]-\widehat{\olep_j^E}-\widehat{\olep_j^C}-\epsilon_j^0$.
The standard error of this expression is $\se \left(  \widehat{\olep_j^{(E \shp C)}} \right) = 
\sd_k[\ln(1-\delta+\delta\exp(\sigma\sqrt{2}u^{(k)}))-\ln(1-\delta+\delta\exp(u^{(k)}-\sigma^2/2))-
\ln(1-\delta+\delta\exp(-\sigma u^{(k)}+\sigma^2/2))]/\sqrt{M}$.
\item Estimate $\olep_j^{[EC]}$, following (\ref{eq:decomp:j:EChard}), as $\widehat{\olep_j^{[EC]}}=\mean_k [-\ln(1-\delta+\delta \exp(\sigma \sqrt{2} u^{(k)}))],$ which has $\se \left( \widehat{\olep_j^{[EC]}} \right) = \sd_k[-\ln(1-\delta+\delta \exp(\sigma \sqrt{2} u^{(k)}))]/\sqrt{M}$.
\item Use $\olep_j^{[E \nat C]}=0$, following (\ref{eq:decomp:j:EnatC}). This is exact.
\item Finally, use $\olr_{j \bs i} = 0$, again exact. 
\end{enumerate}
As for (\ref{eq:decomp_overall_IGR}), it is again easy to see that (\ref{eq:decomp_overall_rjbsi}) applies for the 
estimates as well as for the quantities being estimated.
